The cost of school consolidations: impacts of reduced accessibility

### 2024-04-23 paper/Manuscript.qmd
Update Manuscript.qmd

Accessibility is defined as "access to opportunities" or the "potential for spatial interaction". It is classically presented as the gravity-based measure popularized by @hansenHowAccessibilityShapes1959 and takes the following general multi-modal formulation (Equation (\ref{eq:multi-modal-conventional-accessibility})):
```{=tex}
\begin{equation}
\label{eq:multi-modal-conventional-accessibility}
S_i^m = \sum_{j=1}^JO_j \cdot f^m(c_{ij}^m)
\end{equation}
```
\noindent where:

@@

For the reasons outlined, this paper uses the multimodal spatial availability $V_i^m$ and spatial availability per student $v_i^m$ measures to quantify accessibility changes for the following reasons. First, all residential locations are associated to their respective census DA, TTS TAZ and community boundary (@fig-Fig3) based on spatial location. Each residential location is assigned a number of 'potential' motorized and non-motorized student aged population from these different data sources to calculate a motorized and non-motorized population balancing factor $F_{i}^{pm}$. For each residential location the following is retained: the 2011 and 2016 census average elementary-aged student population per residential location (@fig-Fig2 second row) and the average motorized and non-motorized split as informed by the TTS aggregated by community (@fig-Fig3). Second, the school capacity $O_j$ is estimated and associated with each school (@fig-Fig1). All residential locations are assumed to be able to access all schools by either motorized or non-motorized mode. All origins (residential locations) can reach all schools by motorized mode, but few residential locations can reach schools within a 27 minute non-motorized trip.

Third, the motorized and non-motorized impedance-balancing factors $F_{ij}^{cm}$ are calculated using the modal theoretical TLDs (@fig-Fig4) based on the respective travel times (data source four) for each origin (residential location) to school combination. Finally, outputs from all three stages are joined together. $F_{i}^{pm}$ joined with $F_{ij}^{cm}$ yields the total balancing factor $F^{tm}_{ij}$ which serves to proportionally allocates the school capacity $O_j$ to each residential location. The value is multimodal spatial availability $V_i^m$ which can be interpreted as the number of school-seats that are available to the mode-using population at that residential location. $V_i^m$ is then summed and represented at the DA and Community level along with the calculated $v_i^m$ for interpretation.

@@

The motorized and non-motorized spatial availability $V_i$ for both 2011 and 2016 is represented in the top half of @fig-Fig6. Though at different magnitudes, both populations using non- and motorized modes have higher spatial availability when more proximate to schools. Hence, all DAs have higher values within Hamilton Central and those more proximate to schools in less-urban communities. Trends appear similar for both 2011 and 2016: proximate populations using motorized modes benefit from a wider-range in high spatial availability than non-motorized modes. When $V_im$ is divided by the number of mode-using students in that DA, $v_i^m$ is derived (bottom half of @fig-Fig6). A similar spatial distribution seen in $V_i^m$ can be seen in $v_i^m$ plots, but as a rate of student population. If our aim is to provide sufficient availability to schools for both motorized and non-motorized populations (i.e., greater than 1.0 $v_i^m$ school-seat availability per student), this rate is only achieved in the core of the city for motorized modes, and notably only for non-motorized populations in DAs that are in less-densely populated rural areas that are school proximate and certain pockets of Hamilton Central. The majority of schools closed were within Hamilton Central, so non-motorized populations in DAs within Hamilton Central were the most negatively, and certain motorized populations captured these losses as gains in availability.

@@

To emphasize how spatial availability changed between 2016 and 2011, the change in $V_i^m$ and $v_i^m$ is visualised for both modes in @fig-Fig7. Overall, a decrease in both $V_i^m$ and $v_i^m$ can be seen both modes modes around closed schools (black points). An increase near new or expanded schools (green points) is not as noteable as overall fewer school seats were added relative to population increase. The non-motorized mode-using populations are most impacted by the school closures/consolidations along with other competition factors such as increased children population in less urban areas and an increase in motorized mode use. For non-motorzied mode-using populations, $V_i^m$ decreases across the city. However, some motorized-mode using populations in DAs do experience an increase in $V_i^m$: these DAs are often relatively proximate to schools but more immediately outside walking-range. Under a competitive and constraint calculation, these DAs benefit from the losses in spatial availability that are predominately experienced by the non-motorized populations within Hamilton Central.

@@

# Discussion

@@

Spatial availability can also be summarized by community. The majority of schools were closed in Hamilton Center, this community represents `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA") |> dplyr::select("kid_proportion_2016") |> sum() |> scales::percent()` of the 5-14 year old population in 2016. Though the population between 2011 and 2016 decreased, the rate of school-seat spatial availability decreased disproportionately more (population by `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA") |> transmute(calc = ((sum(kid_2016)-sum(kid_2011))/sum(kid_2011)) |> scales::percent()) |> first()` and spatial availability by `r `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA") |> transmute(calc = ((sum(V_im_2016)-sum(V_im_2011))/sum(V_im_2011)) |> scales::percent())  |> first()`). This out of proportion decrease is greatest in Hamilton Central than any other community. This may have grave equity concerns as Hamilton Central has the highest LIM-AT prevalence, with `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA" & mode == "motorized") |> dplyr::select("prev_LIMAT_2016") |> round() |> paste0("%")` of households LIM-AT in 2016, 3.7 times greater than Glanbrook with the lowest level of LIM-AT prevalence. Between 2016 and 2011, Hamilton Center had the lowest levels of emissions per student and saw the highest per student increase in 2016. Hamilton Center had the highest non-motorized trip potential, and it was reduced in part by the school closure policies.

@@

Across the city, those who commute by non-motorized mode suffer from the most dramatic drop in student population and especially spatial availability across the city. In all communities, spatial availability decreases at greater levels than population. Notably, in Dundas (2nd highest LIM-AT), the only school in the community closed leaving the population in the community with no other option but to use motorized modes. Glanbrook (lowest LIM-AT), which experienced a sharp increase in student population along with non-motorized using population, suffered a lose in non-motorized spatial availability despite seeing the highest increase in motorized spatial availability. Overall, the gain in motorized spatial availability resulted in a further erosion of non-motorized spatial availability. This erosion results in a lose in active transport trips and an increase in emissions city-wide (`r SAvail_im_COMM_changes |> st_drop_geometry() |> transmute(em = (sum(emissions_2011))/(sum(kid_2011)) ) |> sum() |> round(digit=2)`kg CO2e per student in 2011 to `r SAvail_im_COMM_changes |> st_drop_geometry() |> transmute(em = (sum(emissions_2016))/(sum(kid_2016)) ) |> sum() |> round(digit=2)` in 2016).

@@

# Concluding remarks

@@

By quantifying the spatial availability, environmental, and active-travel implications of school closure/consolidation policies in Hamilton, we offer a methodology for spatial policy analysis scenario. This methodology can be used by researchers and decision-makers to plan and evaluate equitable and sustainable urban planning policies from a spatial perspective. We urge policy makers to view 'spatial availability' of public resources from a per-capita perspective, plan capacity and location of service provision with the cost of travel (and associated implications such as GHG emissions, active transportation mode, safety of active travel modes, etc.) that sufficiently serves the target population. In this paper we suggest and assume a benchmark of 1.0 school-seats per student which accounts for sufficy. Spatial availability can be used to obtain this benchmark since the total number of opportunities (i.e., school capacity) is preserved in the region of analysis unlike unconstrained forms of accessibility measurement e.g., Hansen-type measure [@hansenHowAccessibilityShapes1959]. As such, the assigned spatial availability can be divided by population at each zone to yield an interpretable per capita measurement.

>>>>

Accessibility is defined as "access to opportunities" or the "potential for spatial interaction". It is classically presented as the gravity-based measure popularized by @hansenHowAccessibilityShapes1959 and takes the following general multi-modal formulation (Equation (\ref{eq:multi-modal-conventional-accessibility})):

```{=tex}
\begin{equation}
\label{eq:multi-modal-conventional-accessibility}
S_i^m = \sum_{j=1}^JO_j \cdot f^m(c_{ij}^m)
\end{equation}
```
\noindent where:

@@

For the reasons outlined, this paper uses the multimodal spatial availability $V_i^m$ and spatial availability per student $v_i^m$ measures to quantify accessibility changes for the following reasons. First, all residential locations are associated to their respective census DA, TTS TAZ, and community boundary (@fig-Fig3) based on spatial location. Each residential location is assigned a number of 'potential' motorized and non-motorized student aged population from these different data sources to calculate a motorized and non-motorized population balancing factor $F_{i}^{pm}$. For each residential location the following is retained: the 2011 and 2016 census average elementary-aged student population per residential location (@fig-Fig2 second row) and the average motorized and non-motorized split as informed by the TTS aggregated by community (@fig-Fig3). Second, the school capacity $O_j$ is estimated and associated with each school (@fig-Fig1). All residential locations are assumed to be able to access all schools by either motorized or non-motorized mode. All origins (residential locations) can reach all schools by motorized mode, but few residential locations can reach schools within a 27 minute non-motorized trip.

Third, the motorized and non-motorized impedance-balancing factors $F_{ij}^{cm}$ are calculated using the modal theoretical TLDs (@fig-Fig4) based on the respective travel times for each origin (residential location) to school combination. Finally, outputs from all three stages are joined together. $F_{i}^{pm}$ joined with $F_{ij}^{cm}$ yields the total balancing factor $F^{tm}_{ij}$ which serves to proportionally allocate the school capacity $O_j$ to each residential location. The value is multimodal spatial availability $V_i^m$ which can be interpreted as the number of school-seats that are available to the mode-using population at that residential location. $V_i^m$ is then summed and represented at the DA and community level along with the calculated $v_i^m$ for interpretation.

@@

The motorized and non-motorized spatial availability $V_i$ for both 2011 and 2016 is represented in the top half of @fig-Fig6. Though at different magnitudes, both populations using non- and motorized modes have higher spatial availability when more proximate to schools. Hence, all DAs have higher values within Hamilton Central and those more proximate to schools in less-urban communities. Trends appear similar for both 2011 and 2016: proximate populations using motorized modes benefit from a wider-range in high spatial availability than non-motorized modes. When $V_i^m$ is divided by the number of mode-using students in that DA, $v_i^m$ is derived (bottom half of @fig-Fig6). A similar spatial distribution seen in $V_i^m$ can be seen in $v_i^m$ plots, but as a rate of student population. If our aim is to provide sufficient availability to schools for both motorized and non-motorized populations (i.e., greater than 1.0 $v_i^m$ school-seat availability per student), this rate is only achieved in the core of the city for motorized modes, and notably only for non-motorized populations in DAs that are in less-densely populated rural areas that are school proximate and certain pockets of Hamilton Central. The majority of schools closed were within Hamilton Central, so non-motorized populations in DAs within Hamilton Central were the most negatively impacted by losses in availability while certain motorized populations captured these losses as availability gains.

@@

To emphasize how spatial availability changed between 2016 and 2011, the change in $V_i^m$ and $v_i^m$ is visualised for both modes in @fig-Fig7. Overall, a decrease in both $V_i^m$ and $v_i^m$ can be seen across both modes around closed schools (black points). An increase near new or expanded schools (green points) is not as notable as overall fewer school seats were added relative to increases in the school-aged population, leading to increased competition for these resources. For non-motorized mode-using populations, $V_i^m$ decreases across the city. However, some motorized-mode using populations in DAs do experience an increase in $V_i^m$: these DAs are often relatively proximate to schools but more immediately outside walking-range. Under a competitive and constraint-based calculation, these DAs benefit from the losses in spatial availability that are predominately experienced by the non-motorized populations within Hamilton Central.

@@

@@

Spatial availability can also be summarized by community. The majority of schools were closed in Hamilton Center, this community represents `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA") |> dplyr::select("kid_proportion_2016") |> sum() |> scales::percent()` of the 5-14 year old population in 2016. Though the population between 2011 and 2016 decreased, the rate of school-seat spatial availability decreased disproportionately more (population by `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA") |> transmute(calc = ((sum(kid_2016)-sum(kid_2011))/sum(kid_2011)) |> scales::percent()) |> first()` and spatial availability by `r `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA") |> transmute(calc = ((sum(V_im_2016)-sum(V_im_2011))/sum(V_im_2011)) |> scales::percent())  |> first()`). This out of proportion decrease is greatest in Hamilton Central than any other community. This may have equity concerns as Hamilton Central has the highest LIM-AT prevalence, with `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA" & mode == "motorized") |> dplyr::select("prev_LIMAT_2016") |> round() |> paste0("%")` of households LIM-AT in 2016, 3.7 times greater than the community with the lowest level of LIM-AT prevalence in the city. Between 2016 and 2011, Hamilton Center had the lowest levels of emissions per student and saw the highest per student increase in 2016. Hamilton Center had the highest non-motorized trip potential, and it was reduced in part by the school closure policies.

@@

Across the city, those who commute by non-motorized mode see the most dramatic drop in spatial availability across the city. In all communities, spatial availability decreases at greater levels than population. Notably, in Dundas (2nd highest LIM-AT), the only school in the community closed leaving the population in the community with no other option but to use motorized modes. Glanbrook (lowest LIM-AT), which experienced a sharp increase in student population along with non-motorized using population, suffered a lose in non-motorized spatial availability despite seeing the highest increase in motorized spatial availability. Overall, the gain in motorized spatial availability resulted in a further erosion of non-motorized spatial availability. This erosion results in a lose in active transport trips and an increase in emissions city-wide (`r SAvail_im_COMM_changes |> st_drop_geometry() |> transmute(em = (sum(emissions_2011))/(sum(kid_2011)) ) |> sum() |> round(digit=2)`kg CO2e per student in 2011 to `r SAvail_im_COMM_changes |> st_drop_geometry() |> transmute(em = (sum(emissions_2016))/(sum(kid_2016)) ) |> sum() |> round(digit=2)` in 2016).

@@

# Conclusion

@@

By quantifying the spatial availability, environmental, and active-travel implications of school closure/consolidation policies in Hamilton, we offer a methodology for spatial policy analysis scenario. This methodology can be used by researchers and decision-makers to plan and evaluate equitable and sustainable urban planning policies from a spatial perspective. We urge policy makers to view 'spatial availability' of public resources from a per-capita perspective, plan capacity and location of service provision with the cost of travel (and associated implications such as GHG emissions, active transportation mode, safety of active travel modes, etc.) that sufficiently serves the target population. In this paper we suggest and assume a benchmark of 1.0 school-seats per student which accounts for sufficiency. Spatial availability can be used to obtain this benchmark since the total number of opportunities (i.e., school capacity) is preserved in the region of analysis unlike unconstrained forms of accessibility measurement e.g., Hansen-type measure [@hansenHowAccessibilityShapes1959]. As such, the assigned spatial availability can be divided by population at each zone to yield an interpretable per capita measurement.
